### 2021-08-29 README.Rmd
Recreated data package with new foodbank_table.Rmd and ttm_transit.Rmd; knitted README.Rmd after submission to SSM

In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

@@

- Food insecurity  
- Food banks  
- Accessibility  
- COVID-19  
- Hamilton  
- Ontario
- Reproducible research

@@

```{r load-packages, cache=FALSE, include=FALSE}
library(foodbanksHamilton) # Data package
library(kableExtra)
library(patchwork)
library(sf)
library(spdep)
library(tidyverse)
#library(tmap)
```

```{r invoke-data, include=FALSE}
data(data_da_2016)
data(dwelling_network_points_2016)
data(foodbank_locations)
data(foodbank_table)
data(hamilton_boundary)
data(modes_less40k)
data(population_age_income)
data(ttm_car)
data(ttm_transit)
data(ttm_walk)
```

@@

```{r calculate-expected-travel-times, include=FALSE}
## Data preparation
# Calculate expected travel times by mode. These are the travel time by the mode multiplied by the probability of using the mode (which is proxied by the proportion of trips by mode in the DA from the TTS). The total expected travel time is the sum of the expected travel times by mode:
ttm <- ttm %>%
  mutate(e_travel_time_car = travel_time_car * p_car,
         e_travel_time_transit = travel_time_transit * p_transit,
         e_travel_time_walk = travel_time_walk * p_walk,
         e_travel_time = e_travel_time_car + e_travel_time_transit + e_travel_time_walk)
```

\newpage

Introduction
==========================

@@

The emergence of COVID-19, the worst public health crisis since the 1918 flu pandemic, has exposed important social and economic fault lines, and pre-existing patterns of inequality appear to have been exacerbated. Along several other dimensions of stress [e.g., accessibility to health care facilities, @ghorbanzadeh2021spatial; @pereira2021geographic], this seems to be the case for food insecurity as well [@laborde2020poverty]. In the US, for example, it has been estimated that there was an increase of more than 30\% in household food insecurity, and more than one third of households were discovered to be newly food insecure - meaning they did not experience food insecurity before the pandemic [@niles2020early]. In Canada, Men and Tarasuk [-@men2021food] report that about 25\% of individuals who experienced job insecurity (a relatively common occurrence during the pandemic), also experienced food insecurity. Similarly, according to Statistics Canada [-@statisticscanada2020food], in the early stages of the pandemic almost 15\% of individuals reported living in a household that faced food insecurity; the risk of food insecurity was substantially higher for households with children. The difference between households with and without children was significant, and 11.7\% of households with children indicated that "food didn’t last and [there was] no money to get more" sometimes or often, compared to 7.3\% of households without children; likewise, 13\% of households with children indicated that they "[c]ouldn’t afford balanced meals" sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since diet-related diseases, such as obesity, heart-disease, and diabetes, were already critical public health concerns in Canada prior to COVID-19 [@boucher2017ontario]. While food banks are not necessarily a stable solution to food insecurity and in fact may encourage a retrenchment of neoliberal policy [@wakefield2013sweet], at least they can be argued to provide a resource of last instance to households in precarious situations [@bazerghi2016role]. As a mid-size city grappling with deindustrialization, Hamilton exhibits high rates of poverty and food bank use. As recently as 2019, the Hamilton Hunger Report^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf] noted that food banks in the city recorded the highest number of visitors in the past 29 years; a rate of increase greater than population growth. Most troubling, approximately 40% of all visitors were children.

It is known that urban food environments, within which people make their daily food choices, are essential in influencing eating behaviours and health outcomes, based on factors such as food availability, ease of geographic accessibility and socio-demographic variations [@paez2010relative; @vanderlee2017food; @widener2018spatial]. However, while there is a wealth of literature that has examined the topic of geographic accessibility to healthy food through the "food desert" concept, there has been little research into accessibility to food banks. Previous work has explored differences in _accessing_ food banks, such as how some households utilize food banks over short periods of time while others regularly utilize food banks as longer-term resource [e.g., @enns2020experiences]. In addition, transportation and locational considerations have been raised as key issues in food bank accessibility in previous qualitative research [@smith2017food]. However, we are not aware of any research that has focused on estimating or capturing this geographic component of accessibility.

The study of place-based geographic accessibility is concerned with capturing the potential to reach destinations of value using the transportation network [@paez2012measuring]. Indeed, the Government of Canada's recent Food Policy has made "access" to healthy food a priority for Canadian communities^[https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042] and previous survey research has suggested that such accessibility plays a key role in user satisfaction with food bank service delivery [@holmes2018nothing]. However, as with research into the prevalence of food deserts, accessibility to food banks is unlikely to be evenly distributed, and variation throughout a city can be expected due to transportation network characteristics, and the spatial distribution of food bank locations and the population they are meant to serve. Furthermore, policy responses to the COVID-19 pandemic likely have added to the distress of vulnerable households. Non-pharmaceutical interventions during the pandemic involving restrictions in mobility have increased the friction of travel, in particular by transit on which low income populations are more reliant [e.g., @deweese2020tale]. At the same time, the pandemic has created additional stress for the operators of food banks through disruptions in the supply chain [e.g., @mckay2021exploring] as well as concerns surrounding the delivery of service in safe conditions and possible cancellation of food service programs.

For this study, we aim to look at how the landscape of food banks and related services (e.g. low-cost or free meal service providers) available in Hamilton, Ontario, changed during the pandemic. Did the number of open food bank services diminish? If so, what was the accessibility to food banks before the pandemic from the perspective of low income households, and how it changed during the pandemic? And finally, who are most likely to have been impacted by changes in the accessibility landscape? This paper first looks at the distribution of food banks and related services before and during the pandemic. Then, we use the balanced floating catchment area approach of Paez et al. [-@paez2019demand] to investigate the accessibility situation. For this, we adopt a fully disaggregated approach based on parcel-level data. Socio-economic and demographic data are drawn from the latest Census of Canada (2016), whereas travel information is from the most recent regional travel survey from 2016. This paper follows reproducible research recommendations [see @brunsdon2020opening], and the research was conducted using open source tools for transportation analysis [@lovelace2021open]. The code and data necessary to reproduce the analysis are available in a public repository^[https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j_M_BMrXhXFvrV2c?usp=sharing].

Food Insecurity and Food Bank Use in Canada
============

Food insecurity is the inability to acquire and consume an adequate amount or good quality food, leading to inadequate nutrient intake [@enns2020experiences; @kirkpatrick2008food; @tarasuk2009household]. This nutrient deficiency is associated with major population health concerns, particularly among Canadians at socio-economic disadvantage [@bazerghi2016role]. Official government surveys such as the Household Food Security Survey Module (HFSSM), the Canadian Community Health Surveys (CCHS), the Longitudinal and International Study of Adults (LISA), and official classifications determined by Health Canada in relation to socio-demographic variables offer some insight into food insecurity in Canada [@kirkpatrick2008food; @tarasuk2009household; @gundersen2018food]. Nationally, it has been found that food insecurity impacts approximately 12.3% of Canadian households [@tarasuk2014food]

Food banks - sometimes also referred to as ‘food pantries’ and ‘food shelves’ - originated as a community response to aid those with inadequate food by voluntarily offering them meals and ingredients [@riches2002food; @loopstra2012relationship]. The scope and objectives of food banks can vary by region and by country, and these organizations can include not only prepared meals and aliments, but also shared spaces to connect in community gardens and community kitchens [@wakefield2013sweet]. Although in their origin food banks were meant to be provide a temporary solution to accommodate those in hunger due to job retrenchments and economic downfalls since the 1980s, over time they have evolved into a community practice to secure emergency food supplies for those in need [@loopstra2012relationship; @wakefield2013sweet].

In Canada, the number of food banks has steadily increased in the past few decades [@wakefield2013sweet]. The largest database of food banks and their use comes from the non-profit association Food Banks Canada (FBC), which conducts an annual assessment through its affiliated members. FBC's 2018 Hunger Count report^[https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018_FINAL_EN.pdf.aspx?ext=.pdf] (the most recent available) listed 1,830 member food banks across the country, and found that Canadians visited food banks 1.1 million times in March of 2018. Of those accessing food banks, certain population characteristics tend to be over-represented compared to national totals from the 2016 Canadian Census of Population. According to FBC's 2018 data, single-adult households represent 45% of those utilizing food banks despite making up 28% of Canada's population, 19% are single-parent households (compared to 10% nationally), and 35% of those accessing food bank services are children aged 0-18 even though their share of Canada's national population is approximately 20%. In addition, 59% of households accessing food banks list social or disability assistance as their primary source of income. Similarly, using data from the 2011-2012 CCHS, Tarasuk et al. [-@tarasuk2019geographic] found higher odds of food insecurity amongst households relying on social assistance, those without a university degree or with children under the age of 18, and individuals that lived alone, renters, and those identifying as Aboriginal. While surveys revealed that only 20 to 30 percent of those experiencing food insecurity were found to frequent food banks in Canada [@tarasuk2014food], research from Ottawa [@enns2020experiences] and Vancouver [@black2020examining] suggests that long-term users tend to be older, have health or mobility challenges, live in large households, and are less likely to have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects, increases in food insecurity rates due to the onset of COVID-19, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 are not yet available, data from the Daily Bread Food Bank in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city^[https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf], which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

In terms of geography, previous research conducted at the provincial scale using data from the 2011-2012 CCHS found that the prevalence of food insecurity ranged across the country from 11.8% of households in Ontario to 41% of households in Nunavut [@tarasuk2019geographic]. However, although previous research has examined the characteristics of individuals and households accessing food banks, the locational or transportation accessibility aspect of food bank access is not well understood. A wealth of literature examining the food desert concept suggests that, in addition to socio-economic and demographic factors, location and transportation networks play a key role in a household's accessibility to healthy foods [@paez2010relative; @vanderlee2017food; @widener2018spatial]. For food banks specifically, previous qualitative research in Ontario by Smith-Carrier et al. [-@smith2017food] has noted that "transportation can be challenging, particularly if the food bank is situated in a remote location" (p. 32). Particularly, it appears that participants experience challenges with the "inordinate amount of time necessary to obtain food, and difficulties associated with transportation" (p. 39). Users of food banks, according to this research, rely on a variety of modes of transportation to access services. Consequently, the location of facilities matters; in the words of an interviewee: "I wish it [the food bank] was a little more centrally located. Because if I didn’t have a bike I’d have to walk it all the way out there and back. I wonder about people who don’t" (p. 39).

@@

Floating catchment area methods are prone to overestimation of the population and the level of service due to multiple-counting. The population at $P_i$ is allocated to _every_ service point $j$ for which $w_{ij}>0$. Similarly, the level of service at $LOS_j$ is allocated to _every_ population point for which $w_{ji}>0$. This inflation effect has been known for several years, and several modifications have been proposed to mitigate it [e.g., @wan2012three; @delamater2013spatial]. A definitive solution to this issue was presented by Paez et al. [-@paez2019demand]. In order to avoid the multiple-counting in the summations, the population and the level of service need to be allocated _proportionally_. This is achieved by standardizing the weights as follows:
$$
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
$$
\noindent and:
$$
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
$$

The standardized weights satisfy the following conditions:
$$
\sum_{i=1}^nw_{ij}^\text{st}=1
$$

\noindent and:
$$
\sum_{j=1}^Jw_{ji}^\text{st}=1
$$

Since the population is allocated proportionally, its value is preserved:
$$
\sum_{i=1}^nP_iw_{ij}^\text{st}=P_i
$$

\noindent as is the level of service:
$$
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
$$

@@

With a population of around 540,000, the City of Hamilton is the fourth largest city in Ontario. It has historically been home to major manufacturing industries but de-industrialization that has occurred over the past several decades has led Hamilton to become one of the most highly divided cities in Ontario, with a significant proportion of its residents living at or below Canada’s poverty level [@deluca2012code; @jakar2019turning; @latham2007determinants]. In 2016, the Hamilton Community Foundation reported that based on the Low-Income Cut-Off, Hamilton recorded a poverty rate of 16.7%, which was well above the average rate of Ontario (13.7%) and the average national rate (12.8%).^[https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/]. According to data from Hamilton Food Share^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf], approximately 23,000 individuals accessed food banks in the city in March of 2019. Within this total is 9,125 visits by children (minors up to 18 years old), up from 8,278 the year before. Feed Ontario, the province's largest collective of hunger-relief organizations, found that on a per-capita basis, the level of need in the inner core of central Hamilton was second highest in Ontario^[https://feedontario.ca/research/report-hunger-map/].

@@

Data have been prepared for sharing as a data package^[https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j_M_BMrXhXFvrV2c?usp=sharing]. The contents of the data package are described next.

@@

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific a target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. With some providers offering different food services at the same location (e.g. food bank with free and community meal services), and some of these services closing after the onset of the COVID-19 pandemic, we opted to geocode based on the service type. On the other hand, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

```{r table-food-bank-info, echo=FALSE, results='asis'}
foodbank_table %>% kable("html",
                         booktabs = TRUE,
                         align = ("cccccc"),
                         col.names = c("Type", 
                                       "Description", 
                                       "Locations Pre-COVID", 
                                       "Locations During COVID", 
                                       "Additional Notes"),
                         caption = "\\label{tab:food-bank-info}foodbank and Food Service Information.",
                         escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(5, width = "15em")
```

>>>>

In this paper we analyze the changes in accessibility to emergency and community food services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Many of these food services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

@@

- Food insecurity  
- Emergency food services  
- Community food services  
- Food banks  
- Accessibility  
- Balanced Floating Catchment Areas  
- COVID-19  
- Hamilton  
- Ontario
- Reproducible research

@@

```{r load-packages, cache=FALSE, include=FALSE}
library(foodbanksHamilton) # Data package
library(kableExtra)
library(patchwork)
library(sf)
library(spdep)
library(tidyverse)
```

```{r invoke-data, include=FALSE}
data(data_da_2016)
data(dwelling_network_points_2016)
data(foodbank_locations)
data(foodbank_table)
data(hamilton_boundary)
data(modes_less40k)
data(population_age_income)
data(ttm_car)
data(ttm_transit)
data(ttm_walk)

# load updated files manually
#oad("./Accessibility-Foodbanks-Hamilton/output-data-files/ttm_transit.RData")
#load("./Accessibility-Foodbanks-Hamilton/output-data-files/foodbank_table.RData")
```

@@

```{r calculate-expected-travel-times-old, eval = FALSE, include=FALSE}
## Data preparation
# Calculate expected travel times by mode. These are the travel time by the mode multiplied by the probability of using the mode (which is proxied by the proportion of trips by mode in the DA from the TTS). The total expected travel time is the sum of the expected travel times by mode:
#ttm_2021 <- ttm_2021 %>%
#  mutate(e_travel_time_car = travel_time_car * p_car,
#         e_travel_time_transit = travel_time_transit * p_transit,
#         e_travel_time_walk = travel_time_walk * p_walk,
#         e_travel_time = e_travel_time_car + e_travel_time_transit + e_travel_time_walk)
```

```{r calculate-expected-travel-times, include=FALSE}
ttm_long <- ttm %>% transmute(UID, OBJECTID,
                                        car_prob = p_car, 
                                        transit_prob = p_transit, 
                                        walk_prob = p_walk, 
                                        car_traveltime = travel_time_car, 
                                        transit_traveltime = travel_time_transit, 
                                        walk_traveltime = travel_time_walk) %>% 
  pivot_longer(cols = !c("UID", "OBJECTID"), 
               names_to = c("mode", ".value"),
               names_sep = "_") %>%
  group_by(UID, OBJECTID) %>%
  summarize(e_travel_time = weighted.mean(traveltime, prob, na.rm = TRUE))
# income_less40k = first(income_less40k)) %>% 
#  filter(!is.nan(e_travel_time))

ttm <- ttm %>% left_join(ttm_long, by = c("UID", "OBJECTID")) %>% 
  filter(!is.nan(e_travel_time))
```

# Introduction

@@

The emergence of COVID-19, the worst public health crisis since the 1918 flu pandemic, has exposed important social and economic fault lines, and pre-existing patterns of inequality appear to have been exacerbated. Along several other dimensions of stress [e.g., accessibility to health care facilities, @ghorbanzadeh2021spatial; @pereira2021geographic], this seems to be the case for food insecurity as well [@laborde2020poverty]. According to Statistics Canada [-@statisticscanada2020food], in the early stages of the pandemic almost 15% of individuals reported living in a household that faced food insecurity; the risk of food insecurity was substantially higher for households with children. The difference between households with and without children was significant, and 11.7% of households with children indicated that "food didn't last and [there was] no money to get more" sometimes or often, compared to 7.3% of households without children; likewise, 13% of households with children indicated that they "[c]ouldn't afford balanced meals" sometimes or often, compared to 8.8% of households without children. Additionally, Men and Tarasuk [-@men2021food] report that about 25% of individuals who experienced job insecurity (a relatively common occurrence during the pandemic) also experienced food insecurity associated with COVID-related disruptions to employment, financial hardship, and use of food charity.

The impacts of food insecurity during the pandemic are alarming, since diet-related diseases, such as obesity, heart-disease, and diabetes, were already critical public health concerns in Canada prior to COVID-19 [@boucher2017ontario]. While emergency food services are not necessarily a stable solution to food insecurity and in fact may encourage a retrenchment of neoliberal policy [@wakefield2013sweet], in reality provide a resource of last instance to households in precarious situations [@bazerghi2016role]. As a mid-size city grappling with deindustrialization, Hamilton exhibits high rates of poverty and use of emergency food services. As recently as 2019, the Hamilton Hunger Report [@hfs2019] noted that food banks in the city recorded the highest number of visitors in the past 29 years; a rate of increase greater than population growth. Most troubling, approximately 40% of all visitors were children.

It is known that urban food environments, within which people make their daily food choices, are essential in influencing eating behaviours and health outcomes, based on factors such as food availability, ease of geographic accessibility and socio-demographic variations [@paez2010relative; @vanderlee2017food; @widener2018spatial]. However, while there is a wealth of literature that has examined the topic of geographic accessibility to healthy food through the "food desert" concept, there has been little research into accessibility to emergency and community food services. Previous work has explored differences in *accessing* food banks, such as how some households utilize food banks over short periods of time while others regularly utilize food banks as longer-term resource [e.g., @enns2020experiences]. In addition, transportation and locational considerations have been raised as key issues in food bank accessibility in previous qualitative research [@smith2017food]. However, we are not aware of any research that has focused on estimating or capturing this geographic component of accessibility.

The study of place-based geographic accessibility is concerned with capturing the potential to reach destinations of value using the transportation network [@paez2012measuring]. Indeed, the Government of Canada's recent Food Policy [@foodpolicy2019] has made "access" to healthy food a priority for Canadian communities and previous research suggests that such accessibility plays a key role in user satisfaction with food bank service delivery [@holmes2018nothing]. However, as with research into the prevalence of food deserts, accessibility to food banks is unlikely to be evenly distributed, and variation throughout a city can be expected due to transportation network characteristics and the spatial distribution of service locations and the population they are meant to serve. Furthermore, policy responses to the COVID-19 pandemic likely have added to the distress of vulnerable households. Non-pharmaceutical interventions during the pandemic involving restrictions in mobility have increased the friction of travel, in particular by transit on which low income populations are more reliant [e.g., @deweese2020tale]. At the same time, the pandemic has created additional stress for the operators of food banks through disruptions in the supply chain [e.g., @mckay2021exploring] as well as concerns surrounding the delivery of service in safe conditions and possible cancellation of food service programs.

For this study, we aim to look at how the landscape of emergency food and related services (e.g. low-cost or free meal service providers) available in Hamilton, Ontario, changed during the pandemic. Did the number of open services diminish? If so, what was the accessibility to emergency and community food services before the pandemic from the perspective of low income households, and how has it changed during the pandemic with respect to geographic access and congestion at remaining sites? And finally, who are most likely to have been impacted by changes in the accessibility landscape? This paper first looks at the distribution of emergency and community food services before and during the pandemic. Then, we use the balanced floating catchment area approach of Paez et al. [-@paez2019demand] to investigate the accessibility situation. For this, we adopt a fully disaggregated approach based on parcel-level data. Socio-economic and demographic data are drawn from the latest Census of Canada (2016), whereas travel information is from the most recent regional travel survey from 2016. This paper follows reproducible research recommendations [see @brunsdon2020opening], and the research was conducted using open source tools for transportation analysis [@lovelace2021open]. The data and code needed to reproduce the analysis are available in an (anonymous) [Drive Folder](https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j_M_BMrXhXFvrV2c?usp=sharing).

# Food Insecurity and Emergency and Communal Food Services in Canada

Food insecurity is the inability to acquire and consume an adequate amount or good quality food, leading to inadequate nutrient intake [@kirkpatrick2008food] and poorer physical and mental health outcomes [@ramsey2011food; @seligman2010food; @stuff2004household]. In this regard, food insecurity is a major population health concern, particularly among Canadians at socio-economic disadvantage [@bazerghi2016role]. Official government surveys such as the Household Food Security Survey Module (HFSSM), the Canadian Community Health Surveys (CCHS), the Longitudinal and International Study of Adults (LISA), and official classifications determined by Health Canada in relation to socio-demographic variables offer some insight into food insecurity in Canada [@kirkpatrick2008food; @tarasuk2009household; @gundersen2018food].

Nationally, analysis of the 2011-2012 CCHS has previously revealed that food insecurity impacts approximately 12.3% of Canadian households [@tarasuk2014food]. Using the same data, Tarasuk et al. [-@tarasuk2019geographic] found higher odds of food insecurity amongst households relying on social assistance, those without a university degree or with children under the age of 18, and individuals that lived alone, renters, and those identifying as Aboriginal. While surveys revealed that only 20 to 30 percent of those experiencing food insecurity were found to frequent food banks in Canada [@tarasuk2014food], pre-pandemic research from Ottawa [@enns2020experiences] and Vancouver [@black2020examining] suggests that long-term users tend to be older, have health or mobility challenges, live in large households, and are less likely to have employment income. In terms of geography, previous research conducted at the provincial scale using data from the 2011-2012 CCHS found that the prevalence of food insecurity ranged across the country from 11.8% of households in Ontario to 41% of households in Nunavut [@tarasuk2019geographic].

Food banks - sometimes also referred to as 'food pantries' and 'food shelves' - originated as a community response to aid those with inadequate food by voluntarily offering them meals and ingredients [@riches2002food; @loopstra2012relationship]. Although in their origin food banks were meant to provide a temporary solution to accommodate those in hunger due to job retrenchments and economic downfalls since the 1980s, over time many have evolved into a community practice to secure food supplies for those in need [@loopstra2012relationship; @wakefield2013sweet]. In Canada, the number of food banks has steadily increased in the past few decades [@wakefield2013sweet]. The largest database of food banks and their use comes from the non-profit association Food Banks Canada (FBC), which conducts an annual assessment through its affiliated members. FBC's 2018 Hunger Count report [@fbc2018] (the most recent available) listed 1,830 member food banks across the country, and found that Canadians visited food banks 1.1 million times in March of 2018. Of those accessing food banks, certain population characteristics tend to be over-represented compared to national totals from the 2016 Canadian Census of Population. According to FBC's 2018 data, single-adult households represent 45% of those utilizing food banks despite making up 28% of Canada's population, 19% are single-parent households (compared to 10% nationally), and 35% of those using food bank services are children aged 0-18 even though their share of Canada's national population is approximately 20%. In addition, 59% of households accessing food banks list social or disability assistance as their primary source of income.

The COVID-19 pandemic has led to an increase in the number of households living in food insecurity in Canada. Survey results from Statistics Canada from May of 2020 suggest that 14.7% of the population was living in food insecurity in the past 30 days, up from 10.5% in 2017-2018 [@statisticscanada2020food]. Recent data from FBC [@fbc2020] showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 are not yet available, data from the Daily Bread Food Bank [@dbfb2020] in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city, which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

Beyond traditional conceptualizations of food banks as providers of emergency food assistance, other community food services also play an important role in decreasing food insecurity. The scope and objectives of food banks can vary by region and by country, and these organizations can include not only prepared meals and aliments for emergency food supply, but also shared spaces to connect in community gardens and community kitchens [@wakefield2013sweet]. However, the efficacy of these programs in reducing food insecurity differs by the type of service offered. For example, previous qualitative research in the Toronto region has questioned the capacity of community kitchens to improve the food security of low-income households due to their limited scale of operations, and un-subsidized kitchens were found to be particularly inaccessible to to families living in severe poverty [@tarasuk1999qualitative]. However, other food access options such as no-cost or low-cost meals provided through community meals or congregate dining play an important role in decreasing food insecurity. Research in Minnesota found that seniors experiencing food insecurity valued congregate dining for providing affordable meals and a space for social gathering [@oemichen2016investigation]. Furthermore, because seniors paid for the meal, there was no stigma attached to the use of these services compared to traditional food-purchasing assistance such as the Supplemental Nutrition Assistance Program.

While previous research has examined the characteristics of individuals and households accessing emergency and community food services, the locational or transportation accessibility aspect of food bank access is not well understood. A wealth of literature examining the food desert concept suggests that, in addition to socio-economic and demographic factors, location and transportation networks play a key role in a household's accessibility to healthy foods [@paez2010relative; @vanderlee2017food; @widener2018spatial]. For food banks specifically, previous qualitative research in Ontario by Smith-Carrier et al. [-@smith2017food] has noted that "transportation can be challenging, particularly if the food bank is situated in a remote location" (p. 32). Particularly, it appears that participants experience challenges with the "inordinate amount of time necessary to obtain food, and difficulties associated with transportation" (p. 39). Users of food banks, according to this research, rely on a variety of modes of transportation to access services. Consequently, the location of facilities matters; in the words of an interviewee: "I wish it [the food bank] was a little more centrally located. Because if I didn't have a bike I'd have to walk it all the way out there and back. I wonder about people who don't" (p. 39).

@@

Floating catchment area methods are prone to overestimation of the population and the level of service due to multiple-counting. The population at $P_i$ is allocated to *every* service point $j$ for which $w_{ij}>0$. Similarly, the level of service at $LOS_j$ is allocated to *every* population point for which $w_{ji}>0$. This inflation effect has been known for several years, and several modifications have been proposed to mitigate it [e.g., @wan2012three; @delamater2013spatial]. A definitive solution to this issue was presented by Paez et al. [-@paez2019demand]. In order to avoid the multiple-counting in the summations, the population and the level of service need to be allocated *proportionally*. This is achieved by standardizing the weights as follows:

$$
w_{ij}^i = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
$$

\noindent and:

$$
w_{ij}^j = \frac{w_{ij}}{\sum_{j=1}^Jw_{ij}}
$$

The standardized weights satisfy the following conditions:

$$
\sum_{i=1}^nw_{ij}^i=1
$$

\noindent and:

$$
\sum_{j=1}^Jw_{ij}^j=1
$$

Since the population is allocated proportionally, its value is preserved:

$$
\sum_{i=1}^nP_iw_{ij}^i=P_i
$$

\noindent as is the level of service:

$$
\sum_{j=1}^JL_jw_{ij}^j=L_j
$$

@@

With a population of around 540,000, the City of Hamilton is the fourth largest city in Ontario. It has historically been home to major manufacturing industries but de-industrialization that has occurred over the past several decades has led Hamilton to become one of the most highly divided cities in Ontario, with a significant proportion of its residents living at or below Canada's poverty level [@deluca2012code; @jakar2019turning; @latham2007determinants]. The Hamilton Community Foundation [@hcf2018] reported that based on the Low-Income Cut-Off, Hamilton recorded a poverty rate of 16.7% in 2016, which was well above the average rate of Ontario (13.7%) and the average national rate (12.8%). According to data from Hamilton Food Share [@hfs2019], approximately 23,000 individuals accessed food banks in the city in March of 2019. Within this total is 9,125 visits by children (minors up to 18 years old), up from 8,278 the year before. Feed Ontario, the province's largest collective of hunger-relief organizations, found that on a per-capita basis, the level of need in the inner core of central Hamilton was second highest in Ontario [@feedontario2019].

@@

Data have been prepared for sharing in the form of an open data product [see @arribas2021open] available in an (anonymous) [Drive Folder](https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j_M_BMrXhXFvrV2c?usp=sharing). The contents of the data package are described next.

@@

The locations of emergency and community food services were obtained from the Hamilton Public Library's Food Access Guide [@hpl2021]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. With some providers offering different food services at the same location (e.g. food bank with free and community meal services), and some of these services closing after the onset of the COVID-19 pandemic, we opted to geocode based on the service type. On the other hand, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

```{r table-food-bank-info, echo=FALSE, results='asis'}
foodbank_table %>% kable("html",
                         booktabs = TRUE,
                         align = ("cccccc"),
                         col.names = c("Type", 
                                       "Description", 
                                       "Locations Pre-COVID", 
                                       "Locations During COVID", 
                                       "Additional Notes"),
                         caption = "\\label{tab:food-bank-info}foodbank and Food Service Information.",
                         escape = FALSE) %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(5, width = "15em")
```
